---
title: "Data Visualization"
output: html_document
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

## Introduction

In this tutorial, we will play around with the [John Hopkins CSSE Covid-19 dataset](https://github.com/CSSEGISandData/COVID-19).
We will learn basic data wrangling with `tidyverse` packages and data visualization with `ggplot2`.
There is no pre-requisites for this tutorial, but having a basic knowledge about `R`, `RStudio`, and `Rmarkdown` would be useful.

## Some useful tips
- To render the `.Rmd` file into html document using RStudio IDE, press the _knit_ button or press `Cmd + Shift + K` (MacOS) / `Ctrl + Shift + K` (Windows)
- `%>` [(forward pipe) operator](https://magrittr.tidyverse.org/) is used for passing forward the left-hand side values into expressions that appear on the right-hand side 
- `!!` (bang bang) operator is a syntatic sugar for [unqouting code](https://tidyeval.tidyverse.org/sec-why-how.html#unquoting-code)
- Cheat sheets for quick reference:
    - [R Markdown](https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)
    - [Data wrangling with `dplyr` and `tidyr`](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf)
    - [`ggplot2`](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)


## Download and explore data
```{r data}
data_raw <- read_csv("https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

head(data_raw)

```

## Data wrangling

Like any other data analysis task, first we need to clean and restructure our data.

Here we will use some function from `tidyverse` package to format our data.

_Note_: the `!!` (bang bang) operator is a syntatic sugar for [unqouting code](https://tidyeval.tidyverse.org/sec-why-how.html#unquoting-code)

### Renaming and aggregating data
To do analysis by country, first we need to aggregate data on country's province / state

```{r aggregate}
date_start <- NULL
date_end <- NULL

# Rename column names
data <- NULL

head(data)
```

### Pick countries to highlight
Next we want to highlight some countries with the most number of cases in the last day, and perhaps add some additional countries

```{r top_countries}
# arrange country by total number of cases in the last day 
data <- NULL

# Find top 10 countries with most cases
top_countries <- NULL

# Optional (add additional countries)
top_countries <- NULL

# Subset data to include only selected countries
data_top_countries <- NULL
```

### Transposing data
`ggplot2` implements [_A layered grammar of graphics_](https://vita.had.co.nz/papers/layered-grammar.html) framework for data visualization which requires data values to be mapped onto _aesthetics_ (quantifiable features of the resulting graphic).

To do this, we want to transpose the date variables in our data into a single column so that we can represent our data in an `x,y` coordinate, with `x` being the time and `y` being the number of cases.

```{r data_transpose}
data_long <- NULL
  # reformat date
  # 

head(data_long)
```

### Count day after first N case
As countries may be at different phases of the outbreak at a certain date,
we want to convert the date into _day after first N reported cases_
so our results are more comparable.

```{r count_day}
min_N <- 100

data_plot <- NULL
  # group by country
  #

head(data_plot)
```

## Plotting
It's often helpful to break down the task into small steps, for example:

* What do we want to see?
    * We want to see how the number of cases changes over time - a time-series line / scatter plot would be appropriate
    * We also want to compare between-country difference, so we would need to generate one plot for each country

* What are the plot elements?
    * X axis: Period of time (i.e. number of days after first N cases)
    * Y axis: Number of cases
    * Color: each country will have a different color
    * Label: each country will have a different label at the latest recorded day
    
* Anything else?
    * We know that the case growth is likely to be exponential, so we may want the Y-axis in a log scale
    * We want to label the plot
    * We may want to put some extra icing on our plot cake :)
    
Now that we have a plan, it's time to get back to work!

```{r plot}
ggplot(
  data = NULL,
  mapping = aes(
    x = NULL,
    y = NULL,
    color = NULL,
    label = NULL,
    group = NULL
    )
  ) +
  # create line chart
  NULL +
  # create scatter plot
  NULL +
  # add country label
  NULL +
  # tranform y-axis to log scale
  NULL

```

### Decorating the plot
```{r decorate}
## Take 2
ggplot(
  # cut until day 50
  data = NULL,
  mapping = aes(
    x = NULL,
    y = NULL,
    color = NULL,
    label = NULL,
    group = NULL
    )
  ) +
  # make the line and point smaller
  geom_line(NULL) +
  geom_point(NULL) +
  # add country label
  geom_text(NULL,
            # adjust label position
            hjust = 0, nudge_x = 0.5,
            # make smaller text
            size = 3
            ) +
  # Y-axis
  scale_y_continuous(
    # transform to log base 2 scale
    trans = "log2",
    # adjust breaks for multiplicative value: 1,2,5,10,20,50,100,200,500,...
    breaks = c(c(1,2,5)%o% 10^(0:5)),
    # format label for breaks
    labels = scales::number_format(accuracy = 1)
    ) +
  # X-Axis
  scale_x_continuous(
    # adjust breaks: 0, 10, 20, 30, ...
    breaks = seq(0,100,10)
  ) +
  # Change X-Axis, Y-Axis, Title label
  labs(
    title = "Covid-19 cumulative number of cases",
    y = "Number of cases",
    x = paste("Day after first", min_N, "cases")
  ) +
  # Turn off clipping
  coord_cartesian(clip = "off") +
  # change overall theme and font size
  theme_minimal(base_size = 10) +
  # change background 
  theme(
    # hide legend
    NULL,
    # adjust plot margin
    NULL,
    # adjust title
    NULL
  )

  
```

### Extra touch
```{r extra}
library(hrbrthemes)
library(ggrepel)
library(ggthemes)

ggplot(
  # cut until day 50
  data = NULL,
  mapping = aes(
    x = NULL,
    y = NULL,
    color = NULL,
    label = NULL,
    group = NULL,
    fill = NULL)
  ) +
  # make the line and point smaller
  geom_line(NULL) +
  geom_point(NULL) +
  # add country label
  geom_label_repel(
    data = NULL,
    # adjust label position
    hjust = 0, nudge_x = 0.5,
    colour = "white", alpha = 0.95,
    segment.size = 0.2, segment.color = "gray20",
    # make smaller text
    size = 3 ) +
  # highlight last day
  geom_point(
    data = NULL,
    shape = 23, size = 1.1) +
  # change colour
  scale_colour_tableau(palette = "Tableau 20") +
  scale_fill_tableau(palette = "Tableau 20") +
  # Y-axis
  scale_y_continuous(
    trans = "log2",
    # adjust breaks
    breaks = c(c(1,2,5)%o% 10^(0:5)),
    # format breaks
    labels = scales::number_format(accuracy = 1),
    ) +
  # X-Axis
  scale_x_continuous(
    breaks = seq(0,100,10)
  ) +
  # Change X-Axis, Y-Axis, Title label
  labs(
    title = "Covid-19 cumulative number of cases",
    y = "Number of cases",
    x = paste("Day after first", min_N, "cases")
  ) +
  # Turn off clipping
  coord_cartesian(clip = "off") +
  # change overall looks
  theme_ipsum(base_size = 10) +
  # change background 
  theme(
    # hide legend
    legend.position = "none",
    # adjust plot margin
    plot.margin = margin(t=5, r=20, b= 5, l =5),
    # remove minor grid
    panel.grid.minor = element_blank(),
    # adjust title and axis
    text = element_text(color = "grey23"),
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold"),
    # Change background color
    plot.background = element_rect(fill = "#fff1e5")
  )

```

### Saving plots
```{r save}
# by default ggsave will save the last generated plot
# ggsave("my_beautiful_plot.png",
#        width = 8,
#        height = 5)
```
